---
title: "phys-clust"
author: "Youngdeok Hwang"
date: "7/6/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

`bootstrap.test.sub` gives a single permutation results.
`k1` and `k2` are two vectors with estimated $K$'s.

```{r}
# function collections

bootstrap.test.sub <- function(k1,k2){ # this gives a single permutation results
	# k1 and k2 are two vectors with K's
	combined.k <- c(k1,k2)
	k.length <- length(combined.k)
	shuffled <- sample(combined.k)
	k1.b <- shuffled[1:length(k1)]
	k2.b <- shuffled[(1+length(k1)):k.length]
	mean(abs(k1.b - k2.b))
}
```

`bootstrap.test` conducts a bootstrap test. 
It repeats `bootstrap.test.sub` B times.

```{r}
bootstrap.test <- function(k1,k2, B){
	# bootstrap test; repeats bootstrap.test.sub B times
	re <- replicate(B, bootstrap.test.sub(k1,k2))
	test.stat <- mean(abs(k1-k2))
	p.val <- mean(test.stat <= re)
	return(p.val)
}
```


`rqclass` gives a random partition of an interval; baseline is equal partition, but it jitters the boundaries to make sure the EM can start at multiple locations. `k` is the number of partition, `x` is the vector to be partitioned.
```{r}
rqclass <- function (x, k) 
{ # modified from mclust::qclass
  x <- as.vector(x)
  # eps <- sqrt(.Machine$double.eps) 
  # numerical accuracy problem if scale of x is large, so make tolerance
  # scale dependent
  eps <- sd(x)*sqrt(.Machine$double.eps)
  q <- NA
  n <- k
  while(length(q) < (k+1))
  { n <- n + 1
  	qx <- seq(from = 0, to = 1, length = n)
  	qx <- jitter(qx, amount= 0.5/n  )
  	qx[1] <- 0
  	qx[length(qx)] <- 1
    q <- unique(quantile(x, qx)) 
  }
  if(length(q) > (k+1))
  { dq <- diff(q)
    nr <- length(q)-k-1
    q <- q[-order(dq)[1:nr]]
  }
  q[1] <- min(x) - eps
  q[length(q)] <- max(x) + eps
  cl <- rep(0, length(x))
  for(i in 1:k) 
     { cl[ x >= q[i] & x < q[i+1] ] <- i }
  return(cl)
}
```

`rand.gmm` tries GMM and returns BIC.
```{r}
rand.gmm <- function(x, G, model){
# x: the vector of the (random) coefficients 
# G: number of component
# model: takes value between "E" (Equal variance) or "V" (varying variance)

zz <- mclust::unmap(rqclass(x, G)) # random assignment with jittering from the equal partition
delta <- 1000
ms <- mclust::mstep(modelName=model, data=x, z=zz)
old.param <- ms$paraneters$mean # random initial guess for mean
flag <- TRUE
while (delta >1e-5 & flag ){ #EM step
	
	flag <- sum(is.na(ms$parameters$mean)) == 0
	es <- mclust::estep(modelName=model, data=x, z=zz, parameters=ms$parameters)
	zz <- es$z
	ms <- mclust::mstep(modelName=model, data=x, z=zz)
	new.param <- ms$parameters$mean
	delta <- sum(abs(old.param-new.param))
	old.param <- ms$parameters$mean
	
}
 if (flag){
 ZZ <- ms$z
 Ms <- ms$parameters$mean
 S <- sqrt(ms$parameters$variance$sigmasq)
 p <- ms$parameters$pro

 # for matrix calculation
 X <- matrix(rep(x, length(Ms)), nrow=length(x))
 MM <- matrix(rep(Ms,each=length(x)), nrow=length(x))
 PP <- matrix(rep(p ,each=length(x)), nrow=length(x))

 # distinction between E and V
 # get loglikelihood
 if (!(model %in% c("E", "V"))) {stop("model must be either E or V")}
 if (model=="E"){
 SS <- matrix(rep(S, length(Ms)*length(x)), nrow=length(x))
 }
 if (model=="V"){
 SS <- matrix(rep(S ,each=length(x)), nrow=length(x))
 } 
 
 loglik <- sum(log(rowSums(dnorm(X, MM, SS )*PP)))
 b <- bic(modelName=model, loglik=loglik, n=length(x), d=1, G=G, equalPro=FALSE)
 }else{b <- NA}
 return(b)
}
```

`EV.rep` applies GMM `M` times.
```{r}

EV.rep <- function(x, M=100, G){
	# internal function used in K.out.
	r1 <- replicate(M, rand.gmm(x, G, "V"))
	r2 <- replicate(M, rand.gmm(x, G, "E"))
	V <- max(r1, na.rm=TRUE)
	E <- max(r2, na.rm=TRUE)
	out <- c(E,V, G)
	names(out) <- c("E","V", "G")
	return(out)
}
```


`K.out` returns the *estimated* $K$ by selecting the model with the best BIC.
```{r}
K.out <- function(x, M=100, g.max=9){
	evout <- list()
	for (g in 1:g.max){
		evout[[g]] <- EV.rep(x, 200, g)
	}
	sum.out <- do.call("rbind", evout)
	sum.out <- data.frame(sum.out)
	G <- which(sum.out==max(sum.out[,c("E", "V")]), arr.ind=TRUE)[1]
	md <- c("E", "V")[which(sum.out==max(sum.out[,c("E", "V")]), arr.ind=TRUE)[2]]
	return(list(G=G, model=md))
}
```

`colmax` is a subroutine used in the algorithm. 
```{r}
colmax <- function(x){ apply(x, 2, max, na.rm=TRUE)} # useful function
```